Even-point Multi-loop Unitarity and its Applications: Exponentiation, Anomalies and Evanescence

### 2023-07-07 eventPointTwoLoop.tex
typos+wording
fixed wording around optical theorem and some extra typos

\subsection{On-shell Unitarity methods}\label{sec:genU}
Above we have reviewed the known behavior of one-loop integral bases in $D=4$ spacetime dimensions. In this section we will briefly demonstrate why traditional 4D unitarity methods are so powerful -- and why we must go beyond them to accommodate the multi-loop calculations of interest in this work. 
\paragraph{Integrands from 4D Cuts}
Applying one-loop integral reduction in 4D tells us that any one-loop massless amplitude, for which tadpoles are scaleless, can be expressed in terms of box, triangle, and bubble diagrams, with each integral weighted by dimensionally-dependent functions of kinematics:
\begin{equation}\label{eq:4DUn}
\mathcal{A}^{\text{1-loop}}= \sum_{N=2}^4 C_N(D)I_N^D = \sum_{N=2}^4 C_N(4)I_N^D+ \mathcal{R}
\end{equation}
where the scalar basis-integrals, $I_N^D$, are evaluated in $D=4-2\epsilon$, and $\mathcal{R}$ is a rational term that emerges from series expanding around $\epsilon=(4-D)/2$. To construct integral coefficients, $C_N$, we make use of the Optical Theorem for the $S$-matrix, which states that,
\begin{equation}
S^\dagger S = \mathbb{1} \quad \Rightarrow \quad 2\,\text{Im}(T) = T^\dagger T
\end{equation}
where $S=\mathbb{1}+iT$. Applying relation to \eqn{eq:4DUn} allows us to identify the imaginary part of $\mathcal{A}^{\text{1-loop}}$ with the imaginary parts of each basis integral weighted by the respective 4D coefficient, $C_N(4)$:
\begin{equation}\label{eq:UnMeth1loop}
\text{Im}(\mathcal{A}^{\text{1-loop}}) = \sum_{N=2}^4 C_N(4)\, \text{Im}(I_N^D)
\end{equation}
\dj{Wording.} Application of the optical theorem to constrain on-shell amplitudes is known as the unitarity method \cite{UnitarityMethod, BDKUniarityReview}. In order to isolate each $C_N(4)$ appearing in \eqn{eq:4DUn}, we must employ {generalized unitarity}, which places particular internal legs on-shell. That is, for each internal loop propagator, $1/P_i^2$, that we wish to take on-shell, we make the following replacement inside the integral:
\begin{equation}
\frac{i}{P_i^2 + i\delta} \rightarrow 2\pi \delta^{(+)}(P_i^2)
\end{equation}
This replacement of the internal propagators will select out the requisite imaginary parts needed to determine the integral coefficients. This approach to one-loop amplitude construction was systematized in ref.~\cite{Forde:2007mi}, which demonstrated how to directly extract the integrand coefficients directly from 4D cuts:
\begin{align}
C^{(P_iP_jP_k)}_4 &\sim \boxCut
\\
C^{(P_iP_j)}_3 &\sim \triCut - \sum_{k} C^{(P_iP_jP_k)}_4
\\
C^{(P_i)}_2 &\sim \bubCut- \sum_{j} C^{(P_iP_j)} - \sum_{j,k} C^{(P_iP_jP_k)}_4
\end{align}
where exposed legs are summed over all internal states crossing the cut, and the labels $P_i$ are external momentum scales flowing into the specified cuts. The subtractions account for the overlapping information in the respective $N$-gon cuts. By momentum conservation, the momentum flow in the unlabeled vertex is completely determined by the other $N-1$ vertices. As a concrete example, consider the 4D cut construction of the bubble coefficient for pure Born-Infeld theory:
\begin{align}\label{eq:4DBICut}
\bubCutStates{$-$}{$-$}{$+$}{$+$} &= \sum_{h_i\in \text{states}} \mathcal{M}^{\text{BI}}(1^-, 2^-, l_2^{h_2},l_1^{h_1})\mathcal{M}^{\text{BI}}(-l_1^{\bar{h}_1},-l_2^{\bar{h}_2},3^+,4^+)
\\
&= \langle12\rangle^2[l_2 l_1 ]^2 \langle l_1 l_2\rangle^2 [34]^2
\\
&= [(l_1+l_2)^2]^2 \langle12\rangle^2 [34]^2
\\
&= s_{12}^2\langle12\rangle^2 [34]^2
\end{align}
where in the last step we have made the replacement $(l_1+l_2) = -(k_1+k_2)$ by momentum conservation. Determining the leading behavior of the 1-loop Born-Infled amplitude thus amounts to specifying all possible 4D cuts.

@@

\section{Effective Actions} \label{sec:Actions}
In this section we demonstrate how our basis of higher-derivative four-photon operators in \eqn{eq:basisTensors} can be used to construct quantum effective actions that capture loop-level effects. We will distinguish between position space operators that appear in the Lagrangian, $\mathcal{O}$, and their corresponding matrix elements, $\mathcal{T}$, which are on-shell quantities in momentum space:
\begin{equation}\label{eq:OpsVsTens}
\mathcal{T} \equiv \langle \text{out}| \mathcal{O}|\text{in}\rangle
\end{equation}
Similar to the matrix elements, the operators can be expressed in both $D$-dimensional and 4D representations. For example, we can define $\mathcal{T}^{F^2F^2}$ operators as follows:
\begin{align}
\mathcal{O}^{F^2F^2}_{(2,0)} &\sim (D_\mu F_{\rho\sigma}D^\mu F^{\rho\sigma})^2
\\
\mathcal{O}^{F^2F^2}_{(0,1)} &\sim (D_\mu F_{\rho\sigma}D^\nu F^{\rho\sigma})(D_\nu F_{\rho\sigma}D^\mu F^{\rho\sigma})
\end{align}
where the spacetimes indices run from $1,2,...,D$, and $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$ are operator valued abelian field strengths. These operator valued expressions can be normalized so that for an all outgoing four-photon scattering process, $|\text{out}\rangle = |k_1,k_2,k_3,k_4\rangle$, they are related to the on-shell tensor basis elements defined previously in \eqn{eq:basisTensors} as follows:
\begin{align}
\mathcal{T}^{F^2F^2}_{(2,0)} =\langle \text{out}\big|\mathcal{O}^{F^2F^2}_{(2,0)} \big|\text{in}\rangle
\\
\mathcal{T}^{F^2F^2}_{(0,1)} =\langle \text{out} \big|\mathcal{O}^{F^2F^2}_{(0,1)}\big |\text{in}\rangle
\end{align}
Likewise we can construct field theory operators for the split helicity operators using the spinor form of the field strengths:
\begin{equation}
F_\pm = (F^{\mu\nu} \pm i \tilde{F}^{\mu\nu})\sigma^{\mu\nu}_\pm
\end{equation}
where the dual field strength are defined as $\tilde{F}^{\mu\nu}= \frac{1}{2}\epsilon^{\mu\nu\rho\sigma}F^{\mu\nu}$, and the spin matrices are expressed as $\sigma^{\mu\nu}_\pm = \frac{1}{4} \sigma^{[\mu}_\pm \sigma^{\nu]}_\mp$ in terms of the Pauli matrices as follows:
\begin{equation}
\sigma_\pm^\mu = (\mathbb{1},\pm \vec{\sigma})^\mu\,.
\end{equation}
In the above expressions, the spacetime indices now run from $1,...,4$. Using these explicitly 4D operators, we can similarly define operators that select out particular helicity states and derivative structures:
\begin{align}
\mathcal{O}^{\text{4D,1}}_{(--++)} &\sim (D_\mu F_- D_\nu F_-)(D^\mu F_+ D^\nu F_+)
\\
\mathcal{O}^{\text{4D,2}}_{(--++)} &\sim (D_\mu F_- D^\mu F_-)(D_\nu F_+ D^\nu F_+)
\end{align}
where the trace is over the spinor indices of the Pauli matrices. As defined, there operators will produce the 4D helicity structures defined in \eqn{eq:4D1Tens} and \eqn{eq:4D2Tens} when contracted between the outgoing helicity state, $|\text{out}\rangle = |k_1^-,k_2^-,k_3^+,k_4^+\rangle$, and the incoming vacuum state:
\begin{align}
\mathcal{T}^{\text{4D,1}}_{(--++)} &=\langle \text{out}\big| \mathcal{O}^{\text{4D,1}}_{(--++)}  \big|\text{in}\rangle = (s_{13}^2 +s_{14}^2)\langle 12\rangle^2[34]^2
\\
\mathcal{T}^{\text{4D,2}}_{(--++)} &= \langle \text{out}\big|\mathcal{O}^{\text{4D,2}}_{(--++)}  \big|\text{in}\rangle =s_{12}^2 \langle 12\rangle^2[34]^2
\end{align}
In the effective actions we construct in this section, we will reserve $\mathcal{O}$ for effective actions, and $\mathcal{T}$ for on-shell matrix elements used in the EMU construction. However, since the field theory operators are redundant up to field redefinition and equations of motion, we'll find it most convenient to work in terms of the on-shell matrix elements, and then implicitly define the operators in terms of these on-shell expressions.

@@

\mathcal{O}^{4+} &= 2 (D_\mu F_{\alpha \beta} D^\mu  F^{\alpha \beta})^2 - \eta_{\mu(\nu}\eta_{\rho\sigma)}(D_\mu F_{\alpha \beta} D_\nu  F^{\gamma\delta} D_\rho F_{\beta \gamma } D_\sigma F^{\delta \alpha})

>>>>

\subsection{On-shell Unitarity methods}\label{sec:genU}
Above we have reviewed the known behavior of one-loop integral bases in $D=4$ spacetime dimensions. In this section we will briefly demonstrate why traditional 4D unitarity methods are so powerful -- and why we must go beyond them to accommodate the multi-loop calculations of interest in this work. 
\paragraph{Integrands from 4D Cuts}
Applying one-loop integral reduction in 4D tells us that any one-loop massless amplitude, for which tadpoles are scaleless, can be expressed in terms of box, triangle, and bubble diagrams, with each integral weighted by dimensionally-dependent functions of kinematics:
\begin{equation}\label{eq:4DUn}
\mathcal{A}^{\text{1-loop}}= \sum_{N=2}^4 C_N(D)I_N^D = \sum_{N=2}^4 C_N(4)I_N^D+ \mathcal{R}
\end{equation}
where the scalar basis-integrals, $I_N^D$, are evaluated in $D=4-2\epsilon$, and $\mathcal{R}$ is a rational term that emerges from series expanding around $\epsilon=(4-D)/2$. To construct integral coefficients, $C_N$, we make use of the Optical Theorem for the $S$-matrix, which states that,
\begin{equation}
S^\dagger S = \mathbb{1} \quad \Rightarrow \quad 2\,\text{Im}(T) = T^\dagger T
\end{equation}
where $S=\mathbb{1}+iT$. Applying this relation to \eqn{eq:4DUn} allows us to identify the imaginary part of $\mathcal{A}^{\text{1-loop}}$ with the imaginary parts of each basis integral weighted by the respective 4D coefficient, $C_N(4)$:
\begin{equation}\label{eq:UnMeth1loop}
\text{Im}(\mathcal{A}^{\text{1-loop}}) = \sum_{N=2}^4 C_N(4)\, \text{Im}(I_N^D)
\end{equation}
Using the optical theorem to constrain scattering amplitudes is known as the unitarity method \cite{UnitarityMethod, BDKUniarityReview}. In order to isolate each $C_N(4)$ appearing in \eqn{eq:4DUn}, we must place particular internal legs on-shell. That is, for each internal loop propagator that we wish to take on-shell, $1/P_i^2$, we make the following replacement inside the integral:
\begin{equation}
\frac{i}{P_i^2 + i\delta} \rightarrow 2\pi \delta^{(+)}(P_i^2)
\end{equation}
By replacing internal propagators with delta functions, one can isolate the imaginary parts needed to determine the integral coefficients. This approach of taking individual legs on-shell is known as the method of \textit{generalized unitarity} -- and it can be systematically applied to construct all one-loop amplitudes directly from 4D unitarity cuts \cite{Forde:2007mi}. Using generalized unitarity, the integral coefficients correspond to the following set of iterated cuts:
\begin{align}
C^{(P_iP_jP_k)}_4 &\sim \boxCut
\\
C^{(P_iP_j)}_3 &\sim \triCut - \sum_{k} C^{(P_iP_jP_k)}_4
\\
C^{(P_i)}_2 &\sim \bubCut- \sum_{j} C^{(P_iP_j)} - \sum_{j,k} C^{(P_iP_jP_k)}_4
\end{align}
where exposed legs are summed over all internal states crossing the cut, and the labels $P_i$ are external momentum scales flowing into the specified cuts. The subtractions account for the overlapping information in the respective $N$-gon cuts. By momentum conservation, the momentum flow in the unlabeled vertex is completely determined by the other $N-1$ vertices. As a concrete example, consider the 4D cut construction of the bubble coefficient for pure Born-Infeld theory:
\begin{align}\label{eq:4DBICut}
\bubCutStates{$-$}{$-$}{$+$}{$+$} &= \sum_{h_i\in \text{states}} \mathcal{M}^{\text{BI}}(1^-, 2^-, l_2^{h_2},l_1^{h_1})\mathcal{M}^{\text{BI}}(-l_1^{\bar{h}_1},-l_2^{\bar{h}_2},3^+,4^+)
\\
&= \langle12\rangle^2[l_2 l_1 ]^2 \langle l_1 l_2\rangle^2 [34]^2
\\
&= [(l_1+l_2)^2]^2 \langle12\rangle^2 [34]^2
\\
&= s_{12}^2\langle12\rangle^2 [34]^2
\end{align}
where in the last step we have made the replacement $(l_1+l_2) = -(k_1+k_2)$ by momentum conservation. Determining the leading behavior of the 1-loop Born-Infled amplitude thus amounts to specifying all possible 4D cuts.

@@

\section{Effective Actions} \label{sec:Actions}
In this section we demonstrate how our basis of higher-derivative four-photon operators in \eqn{eq:basisTensors} can be used to construct quantum effective actions that capture loop-level effects. We will distinguish between position space operators that appear in the Lagrangian, $\mathcal{O}$, and their corresponding matrix elements, $\mathcal{T}$, which are on-shell quantities in momentum space:
\begin{equation}\label{eq:OpsVsTens}
\mathcal{T} \equiv \langle \text{out}| \mathcal{O}|\text{in}\rangle
\end{equation}
Similar to the matrix elements, the operators can be expressed in both $D$-dimensional and 4D representations. For example, we can define $\mathcal{T}^{F^2F^2}$ operators as follows:
\begin{align}
\mathcal{O}^{F^2F^2}_{(2,0)} &\sim (D_\mu F_{\rho\sigma}D^\mu F^{\rho\sigma})^2
\\
\mathcal{O}^{F^2F^2}_{(0,1)} &\sim (D_\mu F_{\rho\sigma}D^\nu F^{\rho\sigma})(D_\nu F_{\alpha\beta}D^\mu F^{\alpha\beta})
\end{align}
where the spacetimes indices run from $1,2,...,D$, and $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$ are operator valued abelian field strengths. These operator valued expressions can be normalized so that for an all outgoing four-photon scattering process, $|\text{out}\rangle = |k_1,k_2,k_3,k_4\rangle$, they are related to the on-shell tensor basis elements defined previously in \eqn{eq:basisTensors} as follows:
\begin{align}
\mathcal{T}^{F^2F^2}_{(2,0)} =\langle \text{out}\big|\mathcal{O}^{F^2F^2}_{(2,0)} \big|\text{in}\rangle
\\
\mathcal{T}^{F^2F^2}_{(0,1)} =\langle \text{out} \big|\mathcal{O}^{F^2F^2}_{(0,1)}\big |\text{in}\rangle
\end{align}
Likewise we can construct field theory operators for the split helicity operators using the spinor form of the field strengths:
\begin{equation}
F_\pm = (F^{\mu\nu} \pm i \tilde{F}^{\mu\nu})\sigma^{\mu\nu}_\pm
\end{equation}
where the dual field strength are defined as $\tilde{F}^{\mu\nu}= \frac{1}{2}\epsilon^{\mu\nu\rho\sigma}F^{\mu\nu}$, and the spin matrices are expressed as $\sigma^{\mu\nu}_\pm = \frac{1}{4} \sigma^{[\mu}_\pm \sigma^{\nu]}_\mp$ in terms of the Pauli matrices as follows:
\begin{equation}
\sigma_\pm^\mu = (\mathbb{1},\pm \vec{\sigma})^\mu\,.
\end{equation}
In the above expressions, the spacetime indices now run from $1,...,4$. Using these explicitly 4D operators, we can similarly define operators that select out particular helicity states and derivative structures:
\begin{align}
\mathcal{O}^{\text{4D,1}}_{(--++)} &\sim (D_\mu F_- D_\nu F_-)(D^\mu F_+ D^\nu F_+)
\\
\mathcal{O}^{\text{4D,2}}_{(--++)} &\sim (D_\mu F_- D^\mu F_-)(D_\nu F_+ D^\nu F_+)
\end{align}
where the trace is over the spinor indices of the Pauli matrices. As defined, there operators will produce the 4D helicity structures defined in \eqn{eq:4D1Tens} and \eqn{eq:4D2Tens} when contracted between the outgoing helicity state, $|\text{out}\rangle = |k_1^-,k_2^-,k_3^+,k_4^+\rangle$, and the incoming vacuum state:
\begin{align}
\mathcal{T}^{\text{4D,1}}_{(--++)} &=\langle \text{out}\big| \mathcal{O}^{\text{4D,1}}_{(--++)}  \big|\text{in}\rangle = (s_{13}^2 +s_{14}^2)\langle 12\rangle^2[34]^2
\\
\mathcal{T}^{\text{4D,2}}_{(--++)} &= \langle \text{out}\big|\mathcal{O}^{\text{4D,2}}_{(--++)}  \big|\text{in}\rangle =s_{12}^2 \langle 12\rangle^2[34]^2
\end{align}
In the effective actions we construct in this section, we will reserve $\mathcal{O}$ for effective actions, and $\mathcal{T}$ for on-shell matrix elements used in the EMU construction. However, since the field theory operators are redundant up to field redefinition and equations of motion, we'll find it most convenient to work in terms of the on-shell matrix elements, and then implicitly define the operators in terms of these on-shell expressions.

@@

\mathcal{O}^{4+} &= 2 (D_\mu F_{\alpha \beta} D^\mu  F^{\alpha \beta})^2 - \eta^{\mu(\nu}\eta^{\rho\sigma)}(D_\mu F_{\alpha \beta} D_\nu  F^{\gamma\delta} D_\rho F_{\beta \gamma } D_\sigma F^{\delta \alpha})
